abxD01 Paper

### 2015-06-11 Schubert_abxD01_mBio_2015.Rmd
Small edits

**Antibiotics differentially alter the structure of the microbiota and their colonization resistance to** ***C. difficile.*** We selected a panel of seven antibiotics from six classes with the goal of differentially altering the microbiota and assessing their resistance to *C. difficile* colonization (**Table 1**). Following the cessation of antibiotics, each treatment group was given one day of recovery prior to challenge with *C. difficile* spores. One day post challenge we enumerated the density of *C. difficile* in the animals' feces. We observed reproducibly high levels of *C. difficile* colonization in mice treated with cefoperazone, metronidazole, and streptomycin (**Figures 1 and S1**). We observed the most variance in levels of *C. difficile* colonization in mice treated with ampicillin. None of the mice that received ciprofloxacin were colonized. In addition to administering ciprofloxacin by oral gavage, we provided ciprofloxacin by intraperitoneal injection (10 mg/mL). For both approaches we provided one or two days of recovery from the antibiotic treatment. Regardless of the method, the resulting communities were resistant to *C. difficile* colonization. Only one of six mice receiving vancomycin was colonized with *C. difficile.* We suspected that this was due to residual vancomycin repressing *C. difficile* growth. In fact, two days post *C. difficile* challenge, *C. difficile* bloomed in this treatment group to a median of 9.1x10^7^ (interquartile range 7.6x10^7^–1.1x10^8^) CFU/g feces. Furthermore, given two days of post-vancomycin recovery, there was no delay in *C. difficile* colonization to high levels, and on day one post challenge we observed a median of 3.0x10^7^ (interquartile range 2.6x10^7^–3.6x10^7^, N=4) CFU/g feces. These results suggest that although the gut tissue does not absorb vancomycin, the absence of *C. difficile* in the remaining five vancomycin-treated mice may have been due to residual antibiotics lingering in the environment. Overall, these various antibiotic perturbations provided varying levels of colonization by *C. difficile*, which led us to hypothesize that the resulting communities varied in their composition.

@@

**Allowing recovery of the microbiota restores colonization resistance.** In the experiments we have described thus far, we allowed the gut microbiota to recover for 24 hours before challenging them with *C. difficile*. Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C. difficile* is restored {Reeves, 2011 \#4; Theriot, 2014 \#17}. To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover for an additional five days (**Figure S3**). Among the metronidazole-treated mice, those with extended recovery had a $`r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)`$-fold reduction in colonization (P<`r ifelse(metro_test$p.value<0.001, "0.001", "NA")`; **Figure 3**). In addition, `r sum(metro_delay$CFU==0)` of the `r nrow(metro_delay)` mice given the longer recovery period had no detectable *C. difficile* 24 hours after challenge. We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (**Figure 3**). Most notable among these were two OTUs that affiliated with a member of the *Barnesiella* (OTU 2) and the *Escherichia* (OTU 3). The relative abundance of this *Barnesiella*-affiliated OTU increased with the delay, and the relative abundance of this *Escherichia*-affiliated OTU decreased. Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in *C. difficile* colonization (P=`r round(amp_test$p.value, 2)`). As before, we observed a similar increase and decrease in relative abundances for *Barnesiella* (OTU 2) and *Escherichia* (OTU 3)-affiliated OTUs. However untreated, fully resistant mice harbored significantly lower levels of the *Barnesiella*-affiliated OTU (OTU 2). Rather, untreated mice had high levels of various *Porphyromonadaceae*-affiliated OTUs (**Figure 1**). These findings further confirm the context-dependency of colonization resistance suggested by the results of our titration experiments.

@@

**Correlation analysis reveals potentially protective bacteria.** To identify bacterial taxa that could be associated with resistance or susceptibility to *C. difficile* across the three sets of experiments, we measured the correlation between the relative abundance of each OTU on the day of inoculation with the level of *C. difficile* colonization 24 hours later (**Figure 4**). OTUs associated with providing resistance against *C. difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those with associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`). The *Porphyromonadaceae*-affiliated OTUs (ρ<sub>average</sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs) were consistently associated with low levels of *C. difficile* colonization. Among the three *Proteobacteria*-affiliated OTUs with a significant positive association with *C. difficile* colonization, the strongest was affiliated with the *Escherichia* (OTU 3; ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`). By performing an OTU-based analysis we were able to observe intra-family and genus differences in association with *C. difficile* colonization. For example, the *Lachnospiraceae* have been associated with protection against *C. difficile*. Although within the *Lachnospiraceae* family we observed `r number_to_text[sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)]` OTUs that were associated with low levels of *C. difficile* colonization, `r number_to_text[sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)]` OTU was associated with high levels of *C. difficile*. In addition we observed `r number_to_text[nrow(corr_data[families=="Lactobacillaceae",])]` significantly correlated *Lactobacillus*-affiliated OTUs (family *Lactobacillaceae*), `r number_to_text[sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)]` of which were associated with low levels of *C. difficile* and `r number_to_text[sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)]` was associated with high levels. The broad taxonomic representation of OTUs associated with low levels of *C. difficile* again suggests that a diverse community may be advantageous in preventing *C. difficile* colonization.

@@

Previous attempts to study the role of the gut microbiota in colonization resistance against *C. difficile* infection have utilized a single perturbation to the community. Here, we used seven antibiotics from six classes that were given to mice in varying doses and with varying post-antibiotic recovery periods. The result was a combination of 15 different perturbations and the non-perturbed microbiota, which allowed us to generate distinct community profiles that displayed varying susceptibility to *C. difficile* colonization. Our findings demonstrated that colonization resistance was not driven by individual populations, but by a consortium of organisms. Others have demonstrated that *Barnesiella* or *Lachnospiraceae* are partially protective against *C. difficile* {Buffie, 2012 \#9;Reeves, 2012 \#18}. Although we observed similar results in a subset of our perturbations, by using a large number of perturbations, we were able to demonstrate that a varied collection of populations was important for complete colonization resistance. Thus colonization resistance can be described as an emergent property of the microbiome, in which individual bacterial populations integrated in a community contribute to the overall resistance to *C. difficile* {Novikoff, 1945 \#34}.

@@

**Statistical analysis and modeling.** Complete scripts for regenerating our analysis and this paper are available at the online repository for this study (https://github.com/SchlossLab/Schubert\_AbxD01\_mBio\_2015). Comparisons between the antibiotic-treated communities were made by calculating dissimilarity matrices based on the metric of Yue and Clayton {Yue, 2005 \#36}. To avoid biases due to uneven sampling, the dissimilarity matrices were calculated by rarefying the samples to 1,625 sequences per sample. We then used analysis of molecular variance (AMOVA) to test for differences in community structure using 10,000 permutations {Anderson, 2001 \#37}. OTU-based analyses were performed using R (v.3.1.2). After subsampling the OTU frequency data to 1,625 sequences per sample, OTUs were considered for analysis if their average relative abundance within any treatment group was at least 1% (N=38 OTUs). All OTU-by-OTU comparisons were performed using the Kruskal-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests. Comparison of log (base 10) transformed *C. difficile* CFU/g feces between experimental groups was calculated using the Kruskal-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests. Spearman rank correlation analysis was performed between OTU counts and *C. difficile* CFU/g feces. All P-values were corrected using a Benjamini and Hochberg adjustment with an experiment-wide Type I error rate of 0.05 {Benjamini, 1995 \#39}. Random forest regression models were constructed using the randomForest R package using 10,000 trees {Cutler, 2007 \#31}. To construct each tree, two-thirds of the samples were randomly selected were used to train the model and the other one-third of the samples were sampled with replacement to test the model. The regression was performed using the log (base 10) transformation of the number of CFU/g fecal material as the dependent variable and the 38 OTUs as predictor variables.

>>>>

**Antibiotics differentially alter the structure of the microbiota and their colonization resistance to** ***C. difficile.*** We selected a panel of seven antibiotics from six classes with the goal of differentially altering the microbiota and assessing their resistance to *C. difficile* colonization (**Table 1**). Following the cessation of antibiotics, each treatment group was given one day of recovery prior to challenge with *C. difficile* spores. One day post challenge we enumerated the density of *C. difficile* in the animals' feces. We observed reproducibly high levels of *C. difficile* colonization in mice treated with cefoperazone, metronidazole, and streptomycin (**Figures 1 and S1**). We observed the most variance in levels of *C. difficile* colonization in mice treated with ampicillin. None of the mice that received ciprofloxacin were colonized. In addition to administering ciprofloxacin by oral gavage, we provided ciprofloxacin by intraperitoneal injection (10 mg/mL). For both approaches we provided one or two days of recovery from the antibiotic treatment. Regardless of the method, the resulting communities were resistant to *C. difficile* colonization. Only one of six mice receiving vancomycin was colonized with *C. difficile.* We suspected that this was due to residual vancomycin repressing *C. difficile* growth. In fact, two days post *C. difficile* challenge, *C. difficile* bloomed in this treatment group to a median of 9.1x10^7^ (interquartile range 7.6x10^7^–1.1x10^8^) CFU/g feces. Furthermore, given two days of post-vancomycin recovery, there was no delay in *C. difficile* colonization to high levels, and on day one post challenge we observed a median of 3.0x10^7^ (interquartile range 2.6x10^7^–3.6x10^7^, N=4) CFU/g feces. These results suggest that although the gut tissue does not absorb vancomycin, the absence of *C. difficile* in the remaining five vancomycin-treated mice may have been due to residual antibiotics lingering in the environment. Overall, these antibiotic perturbations provided different levels of colonization by *C. difficile*, which led us to hypothesize that the resulting communities varied in their composition.

@@

**Allowing recovery of the microbiota restores colonization resistance.** In the experiments we have described thus far, we allowed the gut microbiota to recover for 24 hours before challenging them with *C. difficile*. Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C. difficile* is restored {Reeves, 2011 \#4; Theriot, 2014 \#17}. To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover for an additional five days (**Figure S3**). Among the metronidazole-treated mice, those with extended recovery had a $`r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)`$-fold reduction in colonization (P<`r ifelse(metro_test$p.value<0.001, "0.001", "NA")`; **Figure 3**). In addition, `r sum(metro_delay$CFU==0)` of the `r nrow(metro_delay)` mice given the longer recovery period had no detectable *C. difficile* 24 hours after challenge. We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (**Figure 3**). Most notable among these were two OTUs that affiliated with a member of the *Barnesiella* (OTU 2) and the *Escherichia* (OTU 3). The relative abundance of this *Barnesiella*-affiliated OTU increased with the extendend recovery, and the relative abundance of this *Escherichia*-affiliated OTU decreased. Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in *C. difficile* colonization (P=`r round(amp_test$p.value, 2)`). As before, we observed a similar increase and decrease in relative abundances for *Barnesiella* (OTU 2) and *Escherichia* (OTU 3)-affiliated OTUs. However untreated, fully resistant mice harbored significantly lower levels of the *Barnesiella*-affiliated OTU (OTU 2). Rather, untreated mice had high levels of various *Porphyromonadaceae*-affiliated OTUs (**Figure 1**). These findings further confirm the context-dependency of colonization resistance suggested by the results of our titration experiments.

@@

**Correlation analysis reveals potentially protective bacteria.** To identify bacterial taxa that could be associated with resistance or susceptibility to *C. difficile* across the three sets of experiments, we measured the correlation between the relative abundance of each OTU on the day of inoculation with the level of *C. difficile* colonization 24 hours later (**Figure 4**). OTUs associated with providing resistance against *C. difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`). The *Porphyromonadaceae*-affiliated OTUs (ρ<sub>average</sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs) were consistently associated with low levels of *C. difficile* colonization. Among the three *Proteobacteria*-affiliated OTUs with a significant positive association with *C. difficile* colonization, the strongest was affiliated with the *Escherichia* (OTU 3; ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`). By performing an OTU-based analysis we were able to observe intra-family and genus differences in association with *C. difficile* colonization. For example, the *Lachnospiraceae* have been associated with protection against *C. difficile* {Reeves, 2012 #18}. Although within the *Lachnospiraceae* family we observed `r number_to_text[sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)]` OTUs that were associated with low levels of *C. difficile* colonization, `r number_to_text[sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)]` OTU was associated with high levels of *C. difficile*. In addition we observed `r number_to_text[nrow(corr_data[families=="Lactobacillaceae",])]` significantly correlated *Lactobacillus*-affiliated OTUs (family *Lactobacillaceae*), `r number_to_text[sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)]` of which were associated with low levels of *C. difficile* and `r number_to_text[sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)]` was associated with high levels. The broad taxonomic representation of OTUs associated with low levels of *C. difficile* again suggests that a diverse community may be advantageous in preventing *C. difficile* colonization.

@@

Previous attempts to study the role of the gut microbiota in colonization resistance against *C. difficile* infection have utilized a single perturbation to the community. Here, we used seven antibiotics from six classes that were given to mice in varying doses and with varying post-antibiotic recovery periods. The result was a combination of 15 different perturbations and the non-perturbed microbiota, which allowed us to generate distinct community profiles that displayed a range of susceptibilities to *C. difficile* colonization. Our findings demonstrated that colonization resistance was not driven by individual populations, but by a consortium of organisms. Others have demonstrated that *Barnesiella* or *Lachnospiraceae* are partially protective against *C. difficile* {Buffie, 2012 \#9;Reeves, 2012 \#18}. Although we observed similar results in a subset of our perturbations, by using a large number of perturbations, we were able to demonstrate that a varied collection of populations was important for complete colonization resistance. Thus colonization resistance can be described as an emergent property of the microbiome, in which individual bacterial populations integrated in a community contribute to the overall resistance to *C. difficile* {Novikoff, 1945 \#34}.

@@

**Statistical analysis and modeling.** Complete scripts for regenerating our analysis and this paper are available at the online repository for this study (https://github.com/SchlossLab/Schubert\_AbxD01\_mBio\_2015). Comparisons between the antibiotic-treated communities were made by calculating dissimilarity matrices based on the metric of Yue and Clayton {Yue, 2005 \#36}. To avoid biases due to uneven sampling, the dissimilarity matrices were calculated by rarefying the samples to 1,625 sequences per sample. We then used analysis of molecular variance (AMOVA) to test for differences in community structure using 10,000 permutations {Anderson, 2001 \#37}. OTU-based analyses were performed using R (v.3.1.2). After subsampling the OTU frequency data to 1,625 sequences per sample, OTUs were considered for analysis if their average relative abundance within any treatment group was at least 1% (N=38 OTUs). All OTU-by-OTU comparisons were performed using the Kruskal-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests. Comparison of log (base 10) transformed *C. difficile* CFU/g feces between experimental groups was calculated using the Kruskal-Wallis rank sum test followed by pairwise Wilcoxon rank sum tests. Spearman rank correlation analysis was performed between OTU counts and *C. difficile* CFU/g feces. All P-values were corrected using a Benjamini and Hochberg adjustment with an experiment-wide Type I error rate of 0.05 {Benjamini, 1995 \#39}. Random forest regression models were constructed using the randomForest R package using 10,000 trees {Cutler, 2007 \#31}. To construct each tree, two-thirds of the samples were randomly selected to train the model and one-third of the samples were selected to test the model. The regression was performed using the log (base 10) transformation of the number of CFU/g fecal material as the dependent variable and the 38 OTUs as predictor variables.
